---
title: "Lives saved by COVID-19 vaccination analysis - update"
output: pdf_document
---

## Supplementary material  
```{r, warning = FALSE, message = FALSE, comment = FALSE, echo = FALSE, fig.height = 15, fig.width = 11}

  Kosovo.footnote <- "[1] All references to Kosovo in this document should be understood to be in the context of the United Nations Security Council resolution 1244 (1999)."


plot.country.mortality.curves(deaths_age, 
                              vax_clean, 
                              variant_waves, 
                              Country_population, 
                              measure = "Variant rates all ages",
                              to.save = TRUE)

```
Supplementary Figure 1. Percentage of estimated lives saved per vaccine dose per age groups, between weeks 50/2020 and 12/2023, in `r n_countries` countries in the European Region. 


\newpage
Supplementary Table 2: Summary table of each WHO European Region countries or areas data availability and whether they were included in the analysis.   
```{r, warning = FALSE, message = FALSE, comment = FALSE, echo = FALSE}

  Supp.table.2.footnote <- paste0(Kosovo.footnote, "1 Calculated as median start and end weeks of neighbouring CAT.\n2 Data from GISAID.\n3 Although Bosnia and Herzegovina is reporting vaccination data, it is not possible to disaggregate by dose for some weeks.\n4 Due to a change in reporting system this country could not be included in the full analysis.\n5 Cannot be disaggregated by age.\n6 Data was completed after data cut.")

  countries_included_summary <- create.data.summary.table(deaths, vaccine, variants, country_codes, 
                                                          countries_partial_reporting, countries_no_reporting) %>% 
    filter(!Period == "Pre-Omicron") %>% 
    select(-Period) %>% 
    mutate(Country_included = ifelse(CountryName %in% countries_partial_reporting, "Partial", Country_included),
           Country_included = ifelse(Country_included == "Yes", "Full", Country_included),
           Variant_data = ifelse(CountryName %in% c("Albania", "Kosovo", "Republic of Moldova"), 
                                 paste0("Yes^1^"), Variant_data),
           Variant_data = ifelse(CountryName %in% c("Israel", "Switzerland"), paste0("Yes^2^"), Variant_data),
           CountryName = ifelse(CountryName == "Kosovo", paste0(CountryName, "[1]"), CountryName),
           CountryName = ifelse(CountryName == "Bosnia and Herzegovina", paste0(CountryName, "^3^"), CountryName),
           CountryName = ifelse(CountryName == "Norway", paste0(CountryName, "^4^"), CountryName),
           CountryName = ifelse(CountryName == "Türkiye", paste0(CountryName, "^5^"), CountryName),
           Vaccination_data = ifelse(CountryName %in% c("Israel", "Netherlands", "Ukraine", "United Kingdom (Scotland)"), 
                                     paste0(Vaccination_data, "^6^"), Vaccination_data),
           Mortality_data = ifelse(CountryName %in% c("Iceland", "Israel", "Kosovo[1]", "Lithuania", "Romania",
                                                      "United Kingdom (Scotland)"), 
                                     paste0(Mortality_data, "^6^"), Mortality_data))

  Supp.Table.2 <- flextable(countries_included_summary) %>%
     theme_vanilla() %>%
     vline(i = NULL, j = c("Age_group", "Variant_data"), part = "all") %>% 
     add_header_row(values = c("", "Date type", ""), colwidths = c(2, 3, 1)) %>%
     set_header_labels(CountryName = "Countries, areas and territories (CAT)",
                       Age_group = "Age group",
                       Mortality_data = "Mortality", 
                       Vaccination_data = "Vaccination", 
                       Variant_data = "Variant", 
                       Country_included = "Included in full or partial analysis?") %>% 
     align(align = "center", part = "all") %>%
     fontsize(size = 10, part = "header") %>%
     fontsize(size = 10, part = "body") %>%
     merge_v(j = "CountryName", target = "CountryName") %>% 
     add_footer_row(values = Supp.table.2.footnote, colwidths = 6) %>%
     set_table_properties(layout = "autofit") %>% 
     fontsize(size = 9, part = "footer")  
  
  Supp.Table.2
  #save_as_docx(Supp.Table.2, path = "Outputs/Deaths_averted_supp_table_2.docx")


```

\newpage
Supplementary Table 3: Age group and country breakdown of vaccination uptake (VU), number of deaths observed and averted as well as observed and expected Mortality Rates (MR), up to week 12/2023 in population aged 25 years and over, for the 34 European Region countries with sufficient data.
```{r, warning = FALSE, message = FALSE, comment = FALSE, echo = FALSE}

Supp.table.3.footnote <- paste0("Mortality rates are per 100,000 population.\n", Kosovo.footnote)


  ST3 <- create.table.4(Expected_cases_all_ages, method = "by age older finer") %>% 
    mutate(TargetGroup = str_replace(TargetGroup, "^Age ", ""),
           TargetGroup = ifelse(TargetGroup == "60+", "≥60",
                              ifelse(TargetGroup == "80+", "≥80", TargetGroup))) %>% 
    filter(!TargetGroup == "TotalDoses") %>% 
    mutate(TargetGroup = factor(TargetGroup, levels = c("25-49", "50-59", "≥60", "60-69", "70-79", "≥80"))) %>% 
    arrange(ReportCountry, TargetGroup)
  
  Supp.Table.3 <- flextable(ST3) %>% 
     theme_vanilla() %>%
     vline(i = NULL, j = c("TargetGroup", "pcDoseAdditional3", "TotalAverted", "ExpectedRate"), part = "all") %>% 
     add_header_row(values = c("", "Vaccination Uptake", "Deaths", "Mortality Rate", ""), colwidths = c(2, 4, 2, 2, 1)) %>%
     set_header_labels(ReportCountry = "CAT",
                       TargetGroup = "Age group (years)",
                       pcSecondDose = "VU2", 
                       pcDoseAdditional1 = "VU3",
                       pcDoseAdditional2 = "VU4",
                       pcDoseAdditional3 = "VU5",
                       DeathsObserved = "Reported", 
                       TotalAverted = "Averted", 
                       ObservedRate = "Reported",
                       ExpectedRate = "Expected",
                       pc_change = "% change") %>% 
     align(align = "center", part = "all") %>%
     fontsize(size = 10, part = "header") %>%
     fontsize(size = 10, part = "body") %>%
     merge_v(j = "ReportCountry", target = "ReportCountry") %>% 
     add_footer_row(values = Supp.table.3.footnote, colwidths = 11) %>%
     set_table_properties(layout = "autofit") %>% 
     fontsize(size = 9, part = "footer")  
  
  Supp.Table.3
  save_as_docx(Supp.Table.3, path = "Outputs/WHO_EURO_Lives_saved_COVID_vaccines_Supp_Table_3.docx")


```


Supplementary Table 4. Age group and country breakdown of vaccination uptake (VU), number of deaths observed and averted as well as observed and expected Mortality Rates (MR), up to week 12/2023 in population aged 25 years and over, for the `r n_countries` European Region countries with sufficient data. Note: the overall total does not include the six countries that could only be included in the pre-Omicron period, namely: Albania, Montenegro, Netherlands, Norway and Poland.   
```{r, warning = FALSE, message = FALSE, comment = FALSE, echo = FALSE}
   
Supp.table.4.footnote <- paste0("Mortality rates are per 100,000 population.\n", Kosovo.footnote)

   # Create Table 4 with age breakdowns per country
  full.time.data <-  read.csv("Outputs/Sensitivity_analyses/Expected_cases_finer_older_2023-W12.csv", 
                               header = TRUE, stringsAsFactors = FALSE) %>% 
    filter(!(ReportCountry %in% countries_partial_reporting | ReportCountry %in% countries_no_reporting)) 
  
  Supp.table.4 <- create.table.4(full.time.data, method = "by age older finer variant") %>% 
    mutate(Opc_change = as.character(Opc_change)) 
  

  # Add countries from pre-omicron period - must have been run before hand
  pre.omicron.data <- read.csv("Outputs/Sensitivity_analyses/Expected_cases_pre_omicron_2023-W12.csv", 
                               header = TRUE, stringsAsFactors = FALSE) %>% 
    select(ReportCountry, year_week, TargetGroup, Variant, 
           FirstDose, SecondDose, DoseAdditional1, DoseAdditional2, DoseAdditional3,
           Denominator, DeathsObserved, 
           DeathsAvertedDose1, DeathsAvertedDose2, DeathsAvertedDose3, DeathsAvertedDose4, DeathsAvertedDose5, 
           TotalAverted, DeathsExpected)

  pre.omicron.data.60 <- pre.omicron.data %>% 
    filter(TargetGroup %in% c("Age 60-69", "Age 70-79", "Age 80+", "Age 60+")) %>% 
    mutate(TargetGroup = "Age 60+") %>% 
    group_by(ReportCountry, TargetGroup, Variant, year_week) %>% 
    summarise(FirstDose = sum(FirstDose),
              SecondDose = sum(SecondDose),
              DoseAdditional1 = sum(DoseAdditional1),
              DoseAdditional2 = sum(DoseAdditional2),
              DoseAdditional3 = sum(DoseAdditional3),
              Denominator = sum(unique(Denominator)),
              DeathsObserved = sum(DeathsObserved),
              DeathsAvertedDose1 = sum(DeathsAvertedDose1),
              DeathsAvertedDose2 = sum(DeathsAvertedDose2),
              DeathsAvertedDose3 = sum(DeathsAvertedDose3),
              DeathsAvertedDose4 = sum(DeathsAvertedDose4),
              DeathsAvertedDose5 = sum(DeathsAvertedDose5),
              TotalAverted = sum(TotalAverted),
              DeathsExpected = sum(DeathsExpected)) %>% 
    ungroup() 
  
  pre.omicron.data <- full_join(pre.omicron.data, pre.omicron.data.60) %>% 
    distinct()
  
  

  Supp.table.4.pre.omicron <- create.table.4(pre.omicron.data, method = "by age older finer variant") %>% 
    filter(ReportCountry %in% countries_partial_reporting)
  

  
  temp.supp.table.4 <- full_join(Supp.table.4, Supp.table.4.pre.omicron) %>% 
     arrange(ReportCountry) %>% 
     mutate(TargetGroup = str_remove(TargetGroup, "Age "),
            TargetGroup = ifelse(TargetGroup == "60+", "≥60",
                              ifelse(TargetGroup == "80+", "≥80", TargetGroup)))

  
Supp.Table.4 <- flextable(temp.supp.table.4) %>%
   theme_vanilla() %>%
   vline(i = NULL, j = c("TargetGroup", "Apc_change", "Dpc_change"), part = "all") %>% 
   vline(i = NULL, j = c("ATotalAverted", "DTotalAverted", "OTotalAverted"), part = "body") %>% 
   add_header_row(values = c("", "Total deaths", "Mortality rates", "Total deaths", "Mortality rates", 
                             "Total deaths", "Mortality rates"),
                  colwidths = c(2, 2, 3, 2, 3, 2, 3)) %>%
   add_header_row(values = c("", "Alpha", "Delta", "Omicron"), colwidths = c(2, 5, 5, 5)) %>%
   set_header_labels(ReportCountry = "CAT",
                     TargetGroup = "Age group\n(years)",
                     ADeathsObserved = "Reported", 
                     ATotalAverted = "Averted", 
                     AObservedRate = "Reported", 
                     AExpectedRate = "Expected", 
                     Apc_change = "% change",
                     DDeathsObserved = "Reported", 
                     DTotalAverted = "Averted", 
                     DObservedRate = "Reported", 
                     DExpectedRate = "Expected", 
                     Dpc_change = "% change",
                     ODeathsObserved = "Reported", 
                     OTotalAverted = "Averted", 
                     OObservedRate = "Reported", 
                     OExpectedRate = "Expected", 
                     Opc_change = "% change") %>% 
   align(align = "center", part = "all") %>%
   fontsize(size = 10, part = "header") %>%
   fontsize(size = 10, part = "body") %>%
   merge_v(j = "ReportCountry", target = "ReportCountry") %>% 
   add_footer_row(values = Supp.table.4.footnote, colwidths = 17) %>%
   set_table_properties(layout = "autofit") %>% 
   fontsize(size = 9, part = "footer")  

Supp.Table.4
save_as_docx(Supp.Table.4, path = "Outputs/WHO_EURO_Lives_saved_COVID_vaccines_Supp_Table_4.docx")


```


\newpage
Supplementary Table 5: Cumulative Vaccination Uptake (VU), as well as observed mortality counts, lives saved per vaccine dose and mortality rates per 100,000 population, by countries, areas and territories (CAT), included in the analysis. Note: d1 refers to first dose, d2, refers to complete series, d3 refers to first additional booster, d4 refers to second additional booster and d5 refers to third additional booster; the ≥60 years age group includes data for all 34 CAT.   
```{r, warning = FALSE, message = FALSE, comment = FALSE, echo = FALSE}

ST5 <- create.table.3(Expected_cases_all_ages, variant_waves, resolution = "Country by age") %>% 
  mutate(Country = ifelse(Country == "Kosovo", "Kosovo[1]", Country),
         Country = ifelse(Country == "Total", "zTotal", Country)) %>% 
  arrange(Country, TargetGroup) %>% 
  mutate(TargetGroup = str_remove(TargetGroup, "Age "),
         TargetGroup = ifelse(TargetGroup == "60+", "≥60",
                              ifelse(TargetGroup == "80+", "≥80", TargetGroup)),
         Country = ifelse(Country == "zTotal", "Total", Country)) 

Supp.Table.5 <- flextable(ST5) %>%
  theme_vanilla() %>%
  vline(i = NULL, j = c("Vaccination uptake Additional 3", "Averted total"), part = "all") %>% 
  add_header_row(values = c("", "Vaccination uptake", "Number of lives", "Mortality rate per 100,000"), 
                 colwidths = c(2, 4, 7, 3)) %>%
  set_header_labels(Country = "Countries, areas and territories",
                    `Age group (years)` = "TargetGroup",
                    `Vaccination uptake Full` = "VU2",
                    `Vaccination uptake Additional 1` = "VU3", 
                    `Vaccination uptake Additional 2` = "VU4", 
                    `Vaccination uptake Additional 3` = "VU5", 
                    `Number of deaths observed` = "Reported deaths",
                    `Averted after one dose` = "Saved after d1", 
                    `Averted after two doses` = "Saved after d2",
                    `Averted after first additional dose` = "Saved after d3", 
                    `Averted after second additional dose` = "Saved after d4",
                    `Averted after third additional dose` = "Saved after d5",
                    `Averted total` = "Total averted", 
                    `Observed mortality rate` = "Reported", 
                    `Total Expected` = "Total Expected",
                    `% expected deaths averted by vaccination` = "% change") %>% 
  align(align = "center", part = "all") %>% 
  merge_v(j = "Country", target = "Country") %>% 
  add_footer_row(values = Kosovo.footnote, colwidths = 16) %>% 
  set_table_properties(layout = "autofit") 


Supp.Table.5
save_as_docx(Supp.Table.5, path = "Outputs/WHO_EURO_Lives_saved_COVID_vaccines_Supp_Table_5.docx")

```


\newpage   
```{r Figure 3, results = "asis", warning = FALSE, message = FALSE, comment = FALSE, echo = FALSE, fig.height = 11, fig.width = 11}

plot.regional.sensitivity.curves(to.save = TRUE)

```
Supplementary Figure 3. Summary of each sensitivity analysis compared to the original scenario and observed mortality, per age-group, in 34 countries in the WHO European Region. Regional and country results are shown in Supplementary Tables 4 and 5. Note: the ≥60 years age group include countries that have only reported ≥60 years data and those reporting by finer age groups (namely, 50-59, 60-69 and ≥80 years).    


\newpage
Supplementary Table 6. Regional summary of sensitivity analyses by age and variant, for all `r n_countries` countries, areas and territories (CAT) included between weeks 50/2020 and 12/2023.
```{r, warning = FALSE, message = FALSE, comment = FALSE, echo = FALSE}

supp.table.6.footnote <- "Scenario descriptions: 1: High VE values; 2: Low VE values; 3: Long lag time; 4: Short lag time; 5: Long vaccine waning time; 6: Short vaccine waning time; 7: High prior immunity; 8: Low prior immunity"

st6 <- create_sensitivity_analyses_summary_table(countries_included, resolution = "Region") %>% 
  mutate(TargetGroup = str_remove(TargetGroup, "Age "),
         TargetGroup = paste0(TargetGroup, " years"),
         TargetGroup = ifelse(TargetGroup == "Total years", "Total", TargetGroup),
         Scenario = substr(Scenario, 1, 1)) %>% 
  arrange(TargetGroup, Scenario)

Supp.Table.6 <- flextable(st6) %>%
  theme_vanilla() %>%
  vline(i = NULL, j = c("Index_pc", "Alpha_pc", "Delta_pc", "Omicron_pc"), part = "all") %>% 
  add_header_row(values = c("", "Index", "Alpha", "Delta", "Omicron", "Total"), 
                 colwidths = c(2, 3, 3, 3, 3, 3)) %>%
  set_header_labels(`TargetGroup` = "Age group",
                    `TotalAverted_Index` = "Total lives\n saved", 
                    `Index_MR` = "Mortality rate\nexpected", 
                    `Index_pc` = "Percentage change",
                    `TotalAverted_Alpha` = "Total lives\n saved", 
                    `Alpha_MR` = "Mortality rate\nexpected", 
                    `Alpha_pc` = "Percentage change",
                    `TotalAverted_Delta` = "Total lives\n saved", 
                    `Delta_MR` = "Mortality rate\nexpected", 
                    `Delta_pc` = "Percentage change",
                    `TotalAverted_Omicron` = "Total lives\n saved", 
                    `Omicron_MR` = "Mortality rate\nexpected", 
                    `Omicron_pc` = "Percentage change",
                    `TotalAverted_Total` = "Total lives\n saved", 
                    `Total_MR` = "Mortality rate\nexpected", 
                    `Total_pc` = "Percentage change") %>% 
  merge_v(j = "TargetGroup", target = "TargetGroup") %>% 
  align(align = "center", part = "all") %>% 
  add_footer_row(values = supp.table.6.footnote, colwidths = 17) %>% 
  set_table_properties(layout = "autofit") 


Supp.Table.6

save_as_docx(Supp.Table.6, path = "Outputs/WHO_EURO_Lives_saved_COVID_vaccines_Supp_Table_6.docx")

```

\newpage
Supplementary table 7. Results of sensitivity analyses according to each alternative Vaccine Effectiveness (VE) scenario, showing the total number of deaths averted for each countries, areas and territories (CAT), expected mortality rate per 100,000 population and the % expected deaths averted by vaccination for the population aged 25 years and older, in `r n_countries` European Region CAT, for weeks 51/2020 to 12/2023.
```{r, warning = FALSE, message = FALSE, comment = FALSE, echo = FALSE}


st7 <- create_sensitivity_analyses_summary_table(countries_included, resolution = "Country") %>% 
  mutate(Scenario_nb = substr(Scenario_nb, 1, 1))

Supp.Table.7 <- flextable(st7) %>%
  theme_vanilla() %>%
  vline(i = NULL, j = c("Scenario_nb", "Deaths_Averted"), part = "all") %>% 
  add_header_row(values = c("", "Deaths", "Mortality rate"), colwidths = c(2, 2, 3)) %>%
  set_header_labels(ReportCountry = "Countries, areas and territories",
                    Scenario_nb = "Sensitivity analysis", 
                    Deaths_Observed = "Reported", 
                    Deaths_Averted = "Averted",
                    Mortality_rate_observed = "Reported", 
                    Mortality_rate_expected = "Expected", 
                    pc_expected = "Percentage change") %>% 
  merge_v(j = "ReportCountry", target = "ReportCountry") %>% 
  add_footer_row(values = supp.table.6.footnote, colwidths = 7) %>% 
  add_footer_row(values = Kosovo.footnote, colwidths = 7) %>% 
  align(align = "center", part = "all") %>% 
  align(align = "left", part = "footer") %>% 
  set_table_properties(layout = "autofit") 


Supp.Table.7

save_as_docx(Supp.Table.7, path = "Outputs/WHO_EURO_Lives_saved_COVID_vaccines_Supp_Table_7.docx")

```
